---
title: "Pneomococcus urinary antigen data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Urinary antigen data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

devtools::load_all()

library(tidyverse)
library(avoncap)
```

```{r setup}

# The data input directory on my development machine:
options("avoncap.input" = "~/Data/avoncap/")

```

# Data loading

The raw data is expected to be in the `urine-antigens` subfolder

```{r}
fs::dir_tree(path = avoncap::input("urine-antigens"))
```

The data is distributed as a single file with both cases and controls together
without distinguishing features. There are BinaxNOW and serotype specific UAD
tests and their results. Not all tests are run for every person.

```{r}
most_recent_files("urine-antigens")
rawUA = load_data("urine-antigens")
```

# Data normalisation

The data can be subdivided into `serotype` versus `binax` groups and normalisation
handles processing the serotype to be in the format `^[0-9]+[A-Z]$`, aligning
format of the patient identifier to match `[0-9]{4}-[0-9]+`. And processing test
dates (which are supplied in the `%e-%b-%y` format).

```{r}

normUA = rawUA %>% normalise_data(instrument = "serotype", .nocache=TRUE)
normBinax = rawUA %>% normalise_data(instrument = "binax", .nocache=TRUE)

```

The normalised UAD data has the individual serotype specific tests and results
in a nested column. The single BinaxNOW result is not nested. Results are keyed
off a surveillance number (admission episode identifier) and test data.

UAD format:

`r avoncap:::.document_data_format(normUA)`

BinaxNOW format:

`r avoncap:::.document_data_format(normBinax)`

# Data augmentation

For the UAD results we commonly want to know the following:

* was the serotype testing complete for all serotypes?
* was the whole UAD panel positive for any serotype, if so how many?
* was the UAD1 panel positive? UAD2?
* was the UAD positive for a PCV vaccine covered serotype?

For the BINAX results there is no real addition qeustions that are relevant


```{r}

augUA = normUA %>% augment_data()


augUA %>% filter(pneumo.serotype_results_complete) %>% 
  group_by(pneumo.serotype_summary_result) %>% 
  count()

augUA %>% unnest(pneumo.pcv_group) %>% with(table(group,result))

ggplot(augUA %>% unnest(pneumo.pcv_group) %>% filter(result == "Positive"), aes(x=group))+geom_histogram(stat="count")



```

# Common operations

```{r}

# normUA %>% unnest(pneumo.urine_antigen) %>% group_by(result) %>% count()
# ggplot(normUA %>% unnest(pneumo.urine_antigen) %>% filter(result == "Positive"), aes(x=test))+geom_histogram(stat="count")
# normBinax %>% group_by(pneumo.binax_result) %>% count()

```

Find all the 

```{r}
custom = bind_rows(
  tibble(group = "PCV7", serotype = uad_groups$pcv7),
  uad_groups$pcv7 %>% .serotype_levels() %>% tibble(group = ., serotype = .),
  tibble(group = "PCV13 (on PCV 7)", serotype = uad_groups$pcv13on7),
  uad_groups$pcv13on7 %>% .serotype_levels() %>% tibble(group = ., serotype = .)
)

tmp = augUA %>% derive_pcv_groupings(pcv_map = custom, col_name = pcv_custom) %>% glimpse()
tmp %>% unnest(pcv_custom) %>% with(table(group,result))

```
